---
title: "Full analysis pipeline"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Full analysis pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 7, fig.height = 5
)
```

This vignette narrates the full analysis pipeline in `analysis/run-all.R`, tracing the connection between scripts and paper results. For a step-by-step walkthrough of the methodology itself, see `vignette("single-iteration")`.

**Prerequisites:** processed data in `data/` and pre-computed results in `rslt/`. Download data from Zenodo with `source("analysis/setup-data.R")`, then reproduce all results with `source("analysis/run-all.R")` (~9 hours total; stability selection and nested validation account for most of that).

```{r libraries, message = FALSE, warning = FALSE}
library(tidyverse)
library(here)
```

## Pipeline overview

`run-all.R` runs five stages in sequence. Data processing requires the unpublished raw sequencing data; all other stages work from the Zenodo-archived processed data.

| Stage | Scripts | Key outputs in `rslt/` |
|-------|---------|------------------------|
| Data processing | `processing/processing-*.R` | `data/*.RData` |
| Stability selection | `stability-selection-*.R` | `stability-selection/*/sel-freq.rds` |
| Nested validation | `nested-validation-*.R` | `stability-selection/*/stable-sets.rds` |
| Model fitting | `model-fitting-*.R` | `models/fitted-models-*.RData` |
| Naive baselines | `naive-preds.R` | `models/naive-preds.rds` |

The line-transect density estimation (`line-transect.R`, see [supplement](https://github.com/ruizt/marine-mammal-edna/blob/main/doc/line-transect.pdf)) is independent of the eDNA modeling pipeline and produces the response variable stored in `data/density-estimates.RData`.

## Stability selection and nested validation

Each `stability-selection-*.R` script fits sPLS across 55 eta × 9 ncomp hyperparameter configurations and 25 LOO subsamples — about 12,375 model fits per species, repeated for three species (~3 hours per marker). Selection frequencies are written to `rslt/stability-selection/<marker>/sel-freq.rds`.

Each `nested-validation-*.R` script then runs an outer LOO loop over the candidate stable sets identified in the previous stage, selecting the candidate that minimizes out-of-sample RMSPE. The winning candidate for each species is written to `rslt/stability-selection/<marker>/stable-sets.rds`.

```{r load-stable-sets}
ss_16s   <- read_rds(here("rslt/stability-selection/16s/stable-sets.rds"))
ss_18sv4 <- read_rds(here("rslt/stability-selection/18sv4/stable-sets.rds"))
ss_18sv9 <- read_rds(here("rslt/stability-selection/18sv9/stable-sets.rds"))

bind_rows(
  mutate(ss_16s,   marker = "16S rRNA"),
  mutate(ss_18sv4, marker = "18S V4"),
  mutate(ss_18sv9, marker = "18S V9")
) |>
  mutate(
    species = recode(species, bm = "Blue", bp = "Fin", mn = "Humpback"),
    n.asv   = map_int(ss, length)
  ) |>
  select(Marker = marker, Species = species,
         ncomp, `Stable ASVs` = n.asv,
         `eta min` = eta.min, `eta max` = eta.max) |>
  arrange(Marker, Species) |>
  knitr::kable(digits = 3, caption = "Stable feature sets selected by nested cross-validation.")
```

The `rmspe.dens` and `cor.dens` columns in `stable-sets.rds` are the nested CV performance estimates for each selected stable set — these are the out-of-sample metrics reported in the paper.

## Model fitting

Each `model-fitting-*.R` script fits a full-data PLS model restricted to the stable feature set chosen for each species. Results are saved to `rslt/models/fitted-models-*.RData`, each containing a `model_metrics` summary and a `fit_df` of observed vs. fitted values.

```{r load-models}
load(here("rslt/models/fitted-models-16s.RData"))
metrics_16s <- model_metrics; fit_df_16s <- fit_df

load(here("rslt/models/fitted-models-18sv4.RData"))
metrics_18sv4 <- model_metrics; fit_df_18sv4 <- fit_df

load(here("rslt/models/fitted-models-18sv9.RData"))
metrics_18sv9 <- model_metrics; fit_df_18sv9 <- fit_df

bind_rows(
  mutate(metrics_16s,   marker = "16S rRNA"),
  mutate(metrics_18sv4, marker = "18S V4"),
  mutate(metrics_18sv9, marker = "18S V9")
) |>
  mutate(species = recode(species, bm = "Blue", bp = "Fin", mn = "Humpback")) |>
  select(Marker = marker, Species = species, `Stable ASVs` = n.asv,
         `Adj. R² (density)` = adj.rsq.dens,
         `Cor. (density)`    = cor.dens,
         `RMSPE (density)`   = rmspe.dens) |>
  arrange(Marker, Species) |>
  knitr::kable(digits = 3, caption = "In-sample model fit. Cor. and RMSPE are from nested LOO cross-validation.")
```

Fitted values vs. observed across all nine marker–species combinations:

```{r pred-obs, fig.cap = "Fitted vs. observed whale density. Dashed line is 1:1."}
bind_rows(
  mutate(fit_df_16s,   marker = "16S rRNA"),
  mutate(fit_df_18sv4, marker = "18S V4"),
  mutate(fit_df_18sv9, marker = "18S V9")
) |>
  mutate(species = recode(species, bm = "Blue", bp = "Fin", mn = "Humpback")) |>
  ggplot(aes(x = dens.obs, y = dens.fit)) +
  geom_point(size = 1.2, alpha = 0.7) +
  geom_abline(linetype = "dashed") +
  facet_grid(species ~ marker, scales = "free") +
  labs(x = "Observed (ind. km\u207b\u00b2)", y = "Fitted (ind. km\u207b\u00b2)")
```

## Naive baselines

`naive-preds.R` computes two species-level baselines for comparison: a lag predictor (previous cruise's density) and a seasonal-mean predictor (mean density for the same calendar quarter). The eDNA RMSPE reported above can be compared against these:

```{r naive}
read_rds(here("rslt/models/naive-preds.rds")) |>
  mutate(species = recode(species, bm = "Blue", bp = "Fin", mn = "Humpback")) |>
  rename(Species = species,
         `RMSPE (lag)`           = rmspe.lag,
         `RMSPE (seasonal mean)` = rmspe.mean) |>
  knitr::kable(digits = 4)
```

## Figures and tables

`analysis/figures.R` and `analysis/tables.R` produce the manuscript figures and supplementary tables from the fitted model objects and density data. `analysis/data.R` exports any derived data objects used elsewhere. These scripts are the last stage in `run-all.R` and require all upstream results to be present in `rslt/`.
