---
title: "Predicting whale density from eDNA: a single-iteration walkthrough"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Predicting whale density from eDNA: a single-iteration walkthrough}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6, fig.height = 4
)
```

This vignette walks through the complete analysis pipeline using one eDNA marker (16S rRNA) and one whale species (blue whale, *Balaenoptera musculus*). The goal is to show the methodology in full — from eDNA feature selection through model fitting to prediction — using a small bundled dataset that requires no additional downloads.

For the complete analysis across all three markers and species, see `vignette("full-analysis")`. For the line-transect density estimation underlying the response variable, see `vignette("line-transect")`.

## Setup

```{r libraries, message = FALSE, warning = FALSE}
library(tidyverse)
library(spls)
library(pls)
library(here)
```

## Data

The example dataset is bundled with the package: a 500-ASV subset of the full 16S data designed so that every ASV identified as stable in the full analysis is present. The companion density estimates are the same 25 observations used in the paper.

```{r load-data}
load(here("data/example-data-16s.RData"))
```

This loads several objects. The two primary inputs are `example_16s` (eDNA data) and `example_density` (blue, fin, and humpback whale log-ratio density estimates):

```{r inspect-data}
dim(example_16s)     # 27 cruises x (1 cruise ID + 500 ASVs)
dim(example_density) # 25 cruises with density estimates for all 3 species
```

Two cruises in `example_16s` have no matching density estimate (due to survey gaps), so the working dataset has 25 observations:

```{r join}
whales_ex <- inner_join(example_density, example_16s, by = "cruise")
cat("n =", nrow(whales_ex), "cruises,  p =",
    sum(startsWith(names(whales_ex), "asv")), "ASVs\n")
```

The response variable — `bm` — is the log-ratio of blue whale density relative to a seasonal mean, specifically $\log(\hat{D} / \bar{D}_\text{season})$ where $\hat{D}$ is the LOO-cross-validated line-transect density estimate (individuals km⁻²) and $\bar{D}_\text{season}$ is the seasonal mean across all cruises. This centering removes gross seasonal trends and focuses the model on finer-scale variation.

```{r response-dist, fig.cap = "Distribution of log-ratio blue whale density across 25 cruises."}
ggplot(whales_ex, aes(x = bm)) +
  geom_histogram(bins = 12) +
  labs(x = "Log-ratio BM density", y = "Count")
```

## Stability selection

The core feature selection method is sparse PLS (sPLS) stability selection (Meinshausen & Bühlmann 2010, adapted for PLS). The idea: fit sPLS across many leave-one-out (LOO) subsamples of the data and a dense hyperparameter grid; features that appear consistently across subsamples are judged "stable."

### Hyperparameter grid

sPLS has two key hyperparameters: `eta` (sparsity, higher = fewer features selected) and `K` (number of PLS components). We search over a grid covering the interesting range:

```{r grid}
eta_vals  <- eta_grid(n = 55)   # 55 values from ~0.4 to ~0.93 on a log scale
ncomp_vals <- 4:12
cat("Grid size:", length(eta_vals), "eta x", length(ncomp_vals), 
    "ncomp =", length(eta_vals) * length(ncomp_vals), "combinations\n")
```

### Running stability selection

For each (eta, ncomp) combination, we fit sPLS to each of the 25 LOO training sets and record which ASVs are selected. The code below shows the full computation; it takes approximately 3 minutes on the 500-ASV example data and is **not evaluated** here — results are loaded from cache in the next step.

```{r stability-selection-code, eval = FALSE}
partitions_ex <- loo_partitions(whales_ex)

model_grid_bm <- expand_grid(
  species = "bm", ncomp = ncomp_vals, eta = eta_vals
)

# ~12,375 sPLS fits; ~3 minutes on the 500-ASV example data
sel_asvs_bm_raw <- pmap(model_grid_bm, function(species, ncomp, eta) {
  fit_spls_partition(species, eta, ncomp, data = whales_ex,
                     partitions = partitions_ex)
}) |> bind_rows()

sel_freq_bm <- selection_frequencies(sel_asvs_bm_raw)

avg_n_bm <- sel_asvs_bm_raw |>
  mutate(n.asv = map_int(sel.asv, length)) |>
  fgroup_by(species, ncomp, eta) |>
  fselect(n.asv) |>
  fmean()
```

Pre-cached versions of `sel_freq_bm`, `avg_n_bm`, and `candidate_sets_bm` were loaded above from `example-data-16s.RData`. See `analysis/create-example-data.R` for the caching code.

> **Note on example vs. full data:** The stability selection false-positive bound is $q_\text{max} = \sqrt{0.6p}$, where $p$ is the total feature count. With $p = 500$ (this example), $q_\text{max} \approx 17$, compared to $q_\text{max} \approx 61$ in the full 16S analysis ($p = 6{,}234$). Stable sets from the example are therefore more conservative (15–25 ASVs vs. 35–48 in the full analysis), but all selected features are confirmed to be genuine stable features from the full analysis. Sensitivity to data perturbations is a known property of stability selection; see the paper for discussion.

### Selection frequencies

For each (eta, ncomp) configuration, we can visualize how consistently each ASV is selected. Here we look at a single configuration near the center of the grid:

```{r sel-freq-plot, fig.cap = "Selection frequency of the 30 most-selected ASVs at ncomp = 4, eta = 0.7. The dashed line marks the pi.max = 0.8 stability threshold."}
sel_freq_bm |>
  filter(ncomp == 4, abs(eta - 0.7) < 0.01) |>
  slice_max(n, n = 30) |>
  mutate(sel.asv = fct_reorder(sel.asv, n),
         freq = n / nrow(whales_ex)) |>
  ggplot(aes(x = freq, y = sel.asv)) +
  geom_col() +
  geom_vline(xintercept = 0.8, linetype = "dashed", colour = "red") +
  labs(x = "Selection frequency", y = NULL,
       title = "ncomp = 4, eta \u2248 0.7")
```

### Stable feature sets

`extract_stable_sets()` scans over all eta intervals and identifies configurations where the average number of selected features ($\bar{q}$) satisfies the false-positive bound. `candidate_sets_bm` contains all such configurations:

```{r candidate-sets}
candidate_sets_bm |> select(species, ncomp, eta.min, eta.max, n.asv)
```

The full analysis uses nested cross-validation (`analysis/nested-validation-16s.R`) to select the best-performing candidate. Here we take the most conservative candidate (narrowest eta interval, implying the tightest false-positive control):

```{r select-stable-set}
best_candidate <- candidate_sets_bm |>
  mutate(interval_width = eta.max - eta.min) |>
  slice_min(interval_width, with_ties = FALSE)

stable_bm <- best_candidate$ss[[1]]
cat("Stable set: ncomp =", best_candidate$ncomp,
    " | eta [", round(best_candidate$eta.min, 3), ",",
    round(best_candidate$eta.max, 3), "]",
    " |", length(stable_bm), "ASVs\n")
```

For reference, `all_stable` contains the 93 ASVs identified as stable for any species in the full 16S analysis. The fraction of the example stable set that is a genuine stable feature:

```{r recovery}
cat(sum(stable_bm %in% all_stable), "/", length(stable_bm),
    "selected ASVs confirmed stable in full analysis\n")
```

## Model fitting

With the stable feature set identified, we fit a standard (non-sparse) PLS regression model using all 25 observations:

```{r model-fit}
data_bm <- subset_to_stable(whales_ex, stable_bm, species = "bm")
fit_bm  <- fit_pls(data_bm, ncomp = best_candidate$ncomp)
```

PLS component loadings describe the linear combination of ASV abundances most associated with whale density. The regression coefficients summarise each feature's net contribution after accounting for all components:

```{r coefficients, fig.cap = "PLS regression coefficients for blue whale density. Positive values indicate ASVs whose abundance increases with blue whale density."}
coefs <- coef(fit_bm)[, , paste(best_candidate$ncomp, "comps")] |>
  enframe(name = "asv", value = "coef") |>
  mutate(direction = if_else(coef > 0, "positive", "negative"))

coefs |>
  slice_max(abs(coef), n = 20) |>
  mutate(asv = fct_reorder(asv, coef)) |>
  ggplot(aes(x = coef, y = asv, fill = direction)) +
  geom_col(show.legend = FALSE) +
  labs(x = "PLS coefficient", y = NULL,
       title = paste("Top 20 coefficients —", length(stable_bm), "ASV model"))
```

### Model metrics

Fitted values on the log-ratio scale and back-transformed predictions on the original density scale:

```{r fitted-values}
lr_obs <- pull(data_bm, y)
lr_fit <- fitted(fit_bm)[, , paste(best_candidate$ncomp, "comps")]

# Back-transform: dens = exp(lr.fit + seasonal_mean)
dens_means_long <- dens_means |>
  pivot_longer(-season, names_to = "species", values_to = "seasonal.mean") |>
  mutate(species = str_remove(species, "log\\.") |> str_remove("\\.imp\\.mean"))

fit_df <- tibble(cruise = whales_ex$cruise, lr.obs = lr_obs, lr.fit = lr_fit) |>
  left_join(dens_raw |> filter(species == "bm") |>
              select(cruise, season, dens.obs = estimate),
            by = "cruise") |>
  left_join(filter(dens_means_long, species == "bm"), by = "season") |>
  mutate(dens.fit  = exp(lr.fit + seasonal.mean),
         lr.resid  = lr.obs - lr.fit,
         dens.resid = dens.obs - dens.fit)

n_obs  <- nrow(fit_df)
ncomp  <- best_candidate$ncomp

adj_rsq_lr   <- 1 - ((n_obs - 1) / (n_obs - ncomp - 1)) *
                    var(fit_df$lr.resid)   / var(fit_df$lr.obs)
adj_rsq_dens <- 1 - ((n_obs - 1) / (n_obs - ncomp - 1)) *
                    var(fit_df$dens.resid) / var(fit_df$dens.obs)

cat(sprintf("Adjusted R\u00b2 (log-ratio scale):  %.3f\n", adj_rsq_lr))
cat(sprintf("Adjusted R\u00b2 (density scale):    %.3f\n", adj_rsq_dens))
```

These figures reflect in-sample fit on all 25 observations; the full analysis reports out-of-sample RMSPE via LOO cross-validation (see `vignette("full-analysis")`).

### Predictions vs. observed

```{r pred-plot, fig.cap = "PLS model fitted values vs. observed blue whale density (individuals km\u207b\u00b2). Points represent individual CalCOFI cruises; the dashed line is the 1:1 reference."}
fit_df |>
  ggplot(aes(x = dens.obs, y = dens.fit)) +
  geom_point() +
  geom_abline(linetype = "dashed") +
  labs(x = "Observed density (ind. km\u207b\u00b2)",
       y = "Fitted density (ind. km\u207b\u00b2)",
       title = "Blue whale — 16S model")
```

## Next steps

- `vignette("full-analysis")` — applies this pipeline to all three markers and three whale species, and compares results across the 9 (marker × species) combinations.
- `vignette("line-transect")` — documents how the observed density estimates used as the response variable were derived from CalCOFI visual survey data.
- `analysis/nested-validation-16s.R` — shows how nested cross-validation selects the best candidate stable set; `analysis/model-fitting-16s.R` — the full-data model fitting for paper reproduction.

## References

Meinshausen, N., & Bühlmann, P. (2010). Stability selection. *Journal of the Royal Statistical Society: Series B*, 72(4), 417–473.

Satterthwaite, E.V., Ruiz, T.D., et al. Microbial and small zooplankton communities predict density of baleen whales in the southern California Current Ecosystem.
