---
title: "Predicting whale density from eDNA: a single-iteration walkthrough"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Predicting whale density from eDNA: a single-iteration walkthrough}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

> **This vignette is a stub.** It demonstrates the full analysis pipeline for a
> single eDNA marker (16S) and a single whale species (blue whale, BM) using a
> bundled 500-ASV example dataset. No additional data downloads are required.
>
> See `vignettes/full-analysis.Rmd` for the complete 3-marker × 3-species
> analysis and `analysis/run-all.R` for paper-exact reproduction.

## Overview

*TODO: narrative introduction to the eDNA → whale density prediction problem*

## Data

```{r load-data}
load(system.file("../data/example-data-16s.RData", package = "marineedna",
                 mustWork = FALSE) |>
       (\(x) if (nchar(x) > 0) x else "data/example-data-16s.RData")())
```

*TODO: describe example_16s, example_density, dimensions, and how the example
data relates to the full dataset*

## Stability selection

*TODO: explain the stability selection concept; show sel_freq_bm; demonstrate
candidate range construction from avg_n_bm; show candidate_sets_bm*

> **Note on computation:** Stability selection on the 500-ASV example data
> (55 η × 9 ncomp × 25 LOO folds = 12,375 sPLS fits) takes approximately
> 3 minutes. Results are pre-cached in `example-data-16s.RData`; the code
> below is shown for illustration but not executed.
>
> The example data contains 500 of the 6,234 ASVs in the full 16S dataset.
> Because the stability selection false-positive bound scales as
> √(0.6 × p), stable sets from the example (p = 500) are smaller (~15–25
> ASVs) than those from the full analysis (~35–48 ASVs). All selected
> features are verified against the full analysis stable set. Sensitivity of
> stable set membership to data perturbations is a known property of the
> method; see Meinshausen & Bühlmann (2010).

## Model fitting

*TODO: use candidate_sets_bm to fit final PLS model; show coefficients*

## Results

*TODO: back-transform predictions using dens_means; compare to observed density
using dens_raw; show adjusted R²*
